\bfseries A Variant of the Sandwich theorem for two-sided limits

### 2025-07-05 lecturenotes_aima104/ch4_aima104.tex
qr factor

\subsection{Exercise}
\begin{enumerate}
 \item Verify parallelogram law: $\Vert u+v \Vert ^2 = 2 \Vert u \Vert ^2 + 2 \Vert v \Vert ^2$.
 \item Suppose $y$ is orthogonal to $u$ and $v$. Show that $y$ is orthogonal to every $w$ in $Span\{u,v\}$.
 \item Let $W$ be a subspace of $\mathbb{R}^n$, then show that $W^{\perp}$ a subspace of $\mathbb{R}^n$.
\item Show that if $x$ is in both $W$ and $W^{\perp}$, then $x=0$.
\end{enumerate}
\subsection{Orthogonal Sets}
A set of vectors $\{ u_1, ...u_p\}$ in a vector space $V$ is said to be \textbf{orthogonal set} if each pair of distinct vectors from the set is orthogonal,i.e for all $u_i, u_j \in V$ we have $u_i.u_j = 0$ whenever $i \neq j$.

\begin{definition}
  An orthogonal basis for a vector space $V$ is a basis for $V$ that is an orthogonal set.
\end{definition}

\begin{theorem}
  \label{orthopro}
  Let $\{u_1, ..., u_p\}$ be an orthogonal basis for a subspace $W$ of $\mathbb{R}^n$. For each $y$ in $W$, the coordinates of $y$ with respect to the orthogonal basis : $y = c_1u_1 + \; ... \; c_n u_n$ are given by $\displaystyle c_j = \frac{y.u_j}{u_j.u_j}$.
\end{theorem}

\subsection{Exercise}
\begin{enumerate}
\item Determine whether the following sets of vectors are orthogonal or not.
  \begin{enumerate}
  \item $
    \begin{bmatrix}
      1 \\ -2 \\ 1
    \end{bmatrix}, \vspace{5mm}
    \begin{bmatrix}
      0 \\ 1 \\ 2
    \end{bmatrix},  \vspace{5mm}
    \begin{bmatrix}
      -5 \\ -2 \\ 1
    \end{bmatrix}$
  \end{enumerate}

@@

  \item Compute the distance of $ y
    \begin{bmatrix}
      3 \\ 1
    \end{bmatrix}
    $ to the line through
    $u= \begin{bmatrix}
      8 \\ 6
    \end{bmatrix}
    $ and the origin.
    \end{enumerate}
\end{enumerate}
\subsection{Orthogonal Projection}
The coordinate $c_j$ of $y$ in Theorem~\ref{orthopro} is actually orthogonal projection of $y$ into the vector $u_j$. This can be generalized. For any given vector $u$. The orthogonal projection of a vector $y$ on $u$ is given by the formula $\displaystyle \hat{y} = \frac{y.u}{u.u}$

\section{Orthonormal Sets}
A set $\{u_1, ..., u_p\}$ is an orthonormal set if it is an orthogonal set of unit vectors. And a basis of orthonormal set is a orthonormal basis. The simplest orthonormal basis is $\{e_1, \; ...\; e_n$ for $\mathbb{R}^n$.

\begin{theorem}
  An $m \times n$ matrix $U$ has orthonormal columns if and only if $U^tU=I$
\end{theorem}

>>>>

\subsection{Exercise}
\begin{enumerate}
 \item Verify parallelogram law: $\Vert u+v \Vert ^2 + \Vert u-v \Vert ^2 = 2 \Vert u \Vert ^2 + 2 \Vert v \Vert ^2$.
 \item Suppose $y$ is orthogonal to $u$ and $v$. Show that $y$ is orthogonal to every $w$ in $Span\{u,v\}$.
 \item Let $W$ be a subspace of $\mathbb{R}^n$, then show that $W^{\perp}$ a subspace of $\mathbb{R}^n$.
\item Show that if $x$ is in both $W$ and $W^{\perp}$, then $x=0$.
\end{enumerate}

\subsection{Orthogonal Sets}
A set of vectors $\{ u_1, ...u_p\}$ in a vector space $V$ is said to be \textbf{orthogonal set} if each pair of distinct vectors from the set is orthogonal,i.e for all $u_i, u_j \in V$ we have $u_i.u_j = 0$ whenever $i \neq j$.
\setcounter{theorem}{2}
\begin{mdframed}
\begin{theorem}
  Any orthogonal set is a linearly independent set.
\end{theorem}
\begin{definition}
  An orthogonal basis for a vector space $V$ is a basis for $V$ that is an orthogonal set.
\end{definition}
\begin{theorem}
  \label{orthopro}
  Let $\{u_1, ..., u_p\}$ be an orthogonal basis for a subspace $W$ of $\mathbb{R}^n$. For each $y$ in $W$, the coordinates of $y$ with respect to the orthogonal basis : $y = c_1u_1 + \; ... \; c_n u_n$ are given by $\displaystyle c_j = \frac{y.u_j}{u_j.u_j}$.
\end{theorem}
\end{mdframed}

\subsection{Orthogonal Projection}
The coordinate $c_j$ of $y$ in Theorem~\ref{orthopro} is actually orthogonal projection of $y$ into the vector $u_j$. This can be generalized. For any given vector $u$. The orthogonal projection of a vector $y$ on $u$ is given by the formula $\displaystyle \hat{y} = \frac{y.u}{u.u}\, u$

\section{Orthonormal Sets, {\mdseries \small \textcolor{red}{Important}}}
\begin{remark}
  A vector having length unity is called unit vector. Given a vector $v$ if we divide it by its magnitude which its length we obtain a unit vector along $v$, i.e $\displaystyle \frac{v}{\Vert v \Vert}$.
\end{remark}

\begin{mdframed}
  A set $\{u_1, ..., u_p\}$ is an orthonormal set if it is an orthogonal set of unit vectors. And a basis of orthonormal set is a orthonormal basis. The simplest orthonormal basis is $\{e_1, \; ...\; e_n$ for $\mathbb{R}^n$.
\end{mdframed}

\subsection{Unitary Matrix}
A real matrix $U$ is said to be unitary if $U^tU=I$. If $U$ is complex then, $\overline{U}\, ^t U= I$, where $\overline{U}\, ^t$ is denoted by ${U}^{\dagger}$.
\begin{theorem}
  An $m \times n$ matrix $U$ has orthonormal columns if and only if $U^tU=I$
\end{theorem}
\subsection{Exercise}
\begin{enumerate}
\item Determine whether the following sets of vectors are orthogonal or not.
  \begin{enumerate}
  \item $
    \begin{bmatrix}
      1 \\ -2 \\ 1
    \end{bmatrix}, \vspace{5mm}
    \begin{bmatrix}
      0 \\ 1 \\ 2
    \end{bmatrix},  \vspace{5mm}
    \begin{bmatrix}
      -5 \\ -2 \\ 1
    \end{bmatrix}$
  \end{enumerate}

@@

  \item Compute the distance of $ y
    \begin{bmatrix}
      3 \\ 1
    \end{bmatrix}
    $ to the line through
    $u= \begin{bmatrix}
      8 \\ 6
    \end{bmatrix}
    $ and the origin.
  \end{enumerate}
\item Take a square unitary matrix of order 2 and a vector $x \in \mathbb{R}^2$ and verify that \\ $\Vert Ux \Vert = \Vert x \Vert$. What can you infer from this property of a unitary matrix?

 \item Show that the Hadamard quantum gate defined by $H =
   \begin{bmatrix}
     1/\sqrt{2} & 1/\sqrt{2} \\[1mm]
     1/\sqrt{2} & -1/\sqrt{2}
   \end{bmatrix}
   $ is a unitary matrix.
 \end{enumerate}

 \section{The Gram-Schmidt Process}
 The Gram-Schmidt Process is a simple algorithm for producing an orthogonal or orthonormal basis for any nonzero subspace of $\mathbb{R}^n$. It uses the concept of \textbf{projection}. First we describe this process when a basis consisting of two vectors $\{x_1,x_2\}$ are given. Now we construct an orthogonal basis $\{v_1, v_2\}$ using the two vectors: $x_1$ and $x_2$ as follows:
 \begin{enumerate}
 \item Let $v_1 = x_1$.
  \item Draw the orthogonal projection from $x_2$ to $v_1$, which is given by $\displaystyle \frac{x_2.v_1}{v_1.v_1} \, v_1$. Now, from the theory of orthogonal projection given above, the vector perpendicular to $v_1$ is given by $\displaystyle v_2 = x_2 - \frac{x_2.v_1}{v_1.v_1} \, v_1$.
 \end{enumerate}

 For a given basis of $p$ vectors $\{x_1,x_2, ..., x_p\}$, we can continue the process of drawing the orthogonal projections and subtraction to get orthogonal vectors as follows:
 %\setcounter{enumerate}{2}
 \begin{enumerate}
 \item To obtain $v_3$ draw the orthogonal projection from $x_3$ onto the Span$v_1,v_2$ and \\[1mm] subtract this from $x_3$: $\displaystyle v_3 = x_3 - \frac{x_3.v_1}{v_1.v_1} \, v_1 - \frac{x_3.v_2}{v_2.v_2} \, v_2$.

  \item $\displaystyle v_p = x_p - \frac{x_p.v_1}{v_1.v_1} \, v_1 - \frac{x_p.v_2}{v_2.v_2} \, v_2 - ... - \frac{x_p.v_{p-1}}{v_1.v_{p-1}} \, v_{p-1}$.
  \end{enumerate}

  \begin{remark}
    After constructing orthogonal basis orthonormal basis can be easily constructed from the orthogonal basis. \textit{How?}
  \end{remark}

  \section{Exercise}
  \begin{enumerate}
  \item Find an orthonormal basis from the given vectors:
    \begin{multicols}{2}
    a). $\begin{bmatrix}
      3 \\ 0 \\ 1
    \end{bmatrix}, \vspace{5mm}
    \begin{bmatrix}
      8 \\ 5 \\ 6
    \end{bmatrix}
    $
    \columnbreak

b).$\begin{bmatrix}
      0 \\ 4 \\ 2
    \end{bmatrix}, \vspace{5mm}
    \begin{bmatrix}
      5 \\ 6 \\ 7
    \end{bmatrix}
    $
  \end{multicols}

\item Find an orthogonal basis for the column space of the matrix:
  \begin{multicols}{2}
    a).  $\begin{bmatrix}
        3 & -5 & 1 \\
        1 & 1 & 1 \\
        -1 & 5 &-2 \\
        3 & -7 & 8
    \end{bmatrix}$
    \columnbreak

    b).
    $\begin{bmatrix}
      -1 & 6 & 6 \\
      3 & -8 & 3 \\
      1 & -2 & 6 \\
      1 & -4 & -3
    \end{bmatrix}$
  \end{multicols}

\item Find the $QR$ factorization of the matrices given in the previous question.
\end{enumerate}

\subsection{QR Factorization}
\begin{theorem}
  If $A$ is an $m \times n$ matrix with linearly independent columns, then $A$ can be factored as $A=QR$, where $Q$ is an $m \times n$ matrix whose columns form an \textbf{orthonormal basis for $Col(A)$} and $R$ is an $n \times n$ \textbf{upper triangular matrix with positive entries on its diagonal}.
\end{theorem}

\begin{proof}
  Given linearly independent columns of the matrix $A$, we can construct $Q$ from the Gram-Schmidt process or any process. Then $Q$ being real unitary matrix we have $Q^t\, Q = I$. Then $Q^t\, A = Q^t\,Q R \implies R = Q^t\, A$.
\end{proof}
\section{Extra Materials}
The angle between the two vectors in plane and space can be generalized by the inner product. For two vectors $u,v \in \mathbb{R}^n$ we have, $u.v = \Vert u \Vert\, \Vert v \Vert \, cos\theta$, where $\theta$ is the angle between the two vectors: $u,v$.

### 2025-07-05 sandwich_theorem/extended_sandwich_theorem.tex
qr factor

Now using this variant Sandwich theorem we can use the above bounds to conclude that \(\displaystyle \lim_{x \to 0} \frac{sinx}{x}=1\) as follows. \\[2mm]
 \(\displaystyle \lim_{x \to 0^-} -\frac{|x|}{x}=1\) \hspace{7cm}
 \(\displaystyle \lim_{x \to 0^+} \frac{|x|}{x}=1\) \\[3mm]

 Hence \(\displaystyle \lim_{x \to 0} \frac{sinx}{x}=1\).

 \section{Remarks on the our variant of Sandwich theorem}
 \begin{itemize}
 \item  While the \(\displaystyle \lim_{x \to 0} \frac{sinx}{x}\) can be determined by the sandwich theorem using the tighter-bound $cosx \leq sinx/x \leq 1$ ~\cite{tao}, the purpose of our variant of sandwich theorem is not to assert that this limit cannot be determined by the Sandwich theorem. This limit can be determined without the sandwich theorem as well.

@@

\item Rudin variant for one-sided limit is analogous to the classical Sandwich theorem for two-side limits. Our variant is not analogous to the classical Sandwich theorem. Rudin's variant has Sandwich theorem for left-hand side limit and for right-hand side limit separately. Our variant has left-hand limit and right-had limit only to loosen the criterion and it is not for the one-sided limit. It is for the two-sided limit.
  \end{itemize}
\vspace{5mm}
\bibliographystyle{aomalpha}
\begin{thebibliography}{9}
\bibitem{hardy}
  Hardy, G. H. (1961) \emph{A Course of Pure Mathematics, 10th edition}. Cambridge University Press, London.
\bibitem{rudin}
  Rudin, W. (1976) \emph{Principles of Mathematical Analysis, 3rd edition}. McGraw Hill, London.
\bibitem{tao}
  Tao, T. (2023) \emph{Analysis I, 2nd edition}. Springer Nature, Singapore.
\bibitem{thomas}
  Thomas, G. B. (2017) \emph{Thomas' Calculus}. Pearson India Education Services Pvt. Ltd, India.
\end{thebibliography}

>>>>

Now using this variant Sandwich theorem we can use the above bounds to conclude that \(\displaystyle \lim_{x \to 0} \frac{sinx}{x}=1\) as follows. \\[2mm]
 \(\displaystyle \lim_{x \to 0^-} -\frac{|x|}{x}=1\) \hspace{7cm}
 \(\displaystyle \lim_{x \to 0^+} \frac{|x|}{x}=1\) \\[3mm]
 Hence \(\displaystyle \lim_{x \to 0} \frac{sinx}{x}=1\).

 \section{Remarks on the our variant of Sandwich theorem}
 \begin{itemize}
 \item  While the \(\displaystyle \lim_{x \to 0} \frac{sinx}{x}\) can be determined by the sandwich theorem using the tighter-bound $cosx \leq sinx/x \leq 1$ ~\cite{boules}, the purpose of our variant of sandwich theorem is not to assert that this limit cannot be determined by the Sandwich theorem. This limit can be determined without the sandwich theorem as well.

@@

\item Rudin variant for one-sided limit is analogous to the classical Sandwich theorem for two-side limits. Our variant is not analogous to the classical Sandwich theorem. Rudin's variant has Sandwich theorem for left-hand side limit and for right-hand side limit separately. Our variant has left-hand limit and right-had limit only to loosen the criterion and it is not for the one-sided limit. It is for the two-sided limit.
  \end{itemize}
\vspace{5mm}
\bibliographystyle{aomalpha}

\begin{thebibliography}{9}
  \bibitem{boules}
  Boules, A. N. (2021). \emph{Fundamentals of Mathematical Analysis}.
\bibitem{hardy}
  Hardy, G. H. (1961) \emph{A Course of Pure Mathematics, 10th edition}. Cambridge University Press, London.
\bibitem{rudin}
  Rudin, W. (1976) \emph{Principles of Mathematical Analysis, 3rd edition}. McGraw Hill, London.
\bibitem{tao}
  Tao, T. (2023) \emph{Analysis I, 2nd edition}. Springer Nature, Singapore.
\bibitem{thomas}
  Thomas, G. B. (2017) \emph{Thomas' Calculus}. Pearson India Education Services Pvt. Ltd, India.

\end{thebibliography}
